# R as a Calculating Environment

The simplest R commands are just mathematical expressions, which are evaluated and the answer is printed out to the console.

```{r}
2 + 2
```

You can write numbers as decimals or using scientific notation, e.g. `0.001` is equivalent to `1e-3`.


## Arithmetic operators

- Addition: `+`
- Subtraction: `-`
- Multiplication: `*`
- Division: `/`
- Exponentiation: `ˆ` (i.e. 'to the power of')
- Integer Division: `%/%` (i.e. division without allowing fractions)
- Modulo: `%%` (i.e. the remainder after integer division)

Note that multiplication/division take place before addition/subtraction. You can change the order of operations using parentheses `()`. e.g.

```{r}
5 + 6 / 2
```

vs.

```{r}
(5 + 6) / 2
```


#### Exercise 2a {-}

1. You have 3 pound coins, 3x 50p, 3x 20p and 4x 5p. Flapjacks cost 89p. How many can you buy?
```{r}
# write the expression below and use Ctrl-Shift-Enter to evaluate it

```

2. How much money will you have left?
```{r}

```


## Mathematical functions and constants

There are a number of basic mathematical functions available, e.g.

- `sqrt()`: square root
- `sin()`, `sinh()` and the other trigonometric / hyperbolic functions
- `asin()`, `asinh()` and the other inverse functions
- `exp()`: natural exponential function
- `log()`, `log10()`: logarithmic functions
- `abs()`: absolute value
- `round()`: round to the closest integer
- `floor()`: round down
- `ceiling()`: round up

Also `pi` for mathematical $\pi$

e.g.
```{r}
cos(pi/3)
```




#### Exercise 2b {-}

What happens when the input provided is outside the domain of a function (e.g. log(-4))?

```{r}

```




## Basic data types in R

R works with numerous data types. Here are some of the most basic ones:

- Decimal values like `4.5` are called `numeric`.
- Natural numbers written like `4L` are called `integer`. Integers are also `numeric`.
- Boolean values (`TRUE` or `FALSE`) are called `logical`.
- Text (or string) values (`"a"`,`"apple"`,`"54"` etc.) are called `character`.
- Complex numbers (e.g. `5+3i`) are called `complex`.

To check the data type of an object, the function `class()` is used. You can also test data types using `is.*()`, e.g.
```{r}
class(154)
class(154L)
is.integer(154)
is.logical(FALSE)
```




## Variables

A **variable** is a basic concept in all programming. You can think of a
variable as a labelled container that allows you to store a value (e.g. 55) or an object (e.g. a data structure).

The assignment operator `<-` is preferred for variable assignment in R, but `=` will also work.

```{r}
a <- 5
b = 9
```

You can then use this variable’s name to easily access the value or the object that is stored within this variable.
Use `print()` or `show()` to display values in the console window. Putting parentheses around an expression is another alternative.

```{r}
print(a)
show(b)
(a+b)
```

You can easily change the value assigned to a variable.
```{r}
a <- 101
print(a)

a <- a + 10
print(a)
```

The variable is retained in memory (see the environment window at top right) until deletion with `rm()`.

```{r}
rm(a)
print(a)
```

Note that rm is *not* reversible.

Variables should have names that are *meaningful*, and should follow the R syntax
rules, that the name should consist of only

- letters (a-z / A-Z)
- numbers
- period `.`
- underscore `_`

A name must start with a letter or period. Names starting with a period are hidden by default.

A common problem is trying to use names that are already in use by R:
e.g. reserved words like `if` or `break` or system objects like `mean`, `TRUE`,
and `pi`.

Another common irritation for new users is that R is *case sensitive*,
so `T` (a system representation of logical true) is different to `t` (function for matrix transpose).



## Logical expressions

The comparison operators known to R are:

- `<` for less than
- `>` for greater than
- `<=` for less than or equal to
- `>=` for greater than or equal to
- `==` for equal to each other
- `!=` not equal to each other

e.g.
```{r}
a <- 21
a > 12
```

The output of a logical expression is one of the *boolean* values `TRUE` or `FALSE`.

We will also need to handle more complex comparisons, like ”greater than 12, and even.” Such comparisons require the *logical operators*:

- `!` logical NOT
- `&` logical AND
- `∣` logical OR

```{r}
a > 12 & a %% 2 == 0
```


## Precision and rounding error

R calculates all values to a high precision, but will only display the first 7 significant figures by default.
You can change this using a command like `options(digits = 10)` (which shows the first 10 digits).

However, only fractions whose denominator is a power of 2 can be represented exactly by R.
All other real values are subject to rounding error, which can cause a lot of confusion when testing for equality, e.g.

```{r}
sqrt(2) * sqrt(2) == 2
```

To get around this issue, we use the function `all.equal()`, which returns `TRUE` if the difference between its two arguments is less than R's level of accuracy:

```{r}
all.equal(sqrt(2)*sqrt(2),2)
```


## Vectors

Vectors are the fundamental data structure in R. A vector is an
*integer-indexed* one-dimensional array that can hold of data of the
*same* type.

```{r}
my.integers <- 1:10 # sequence of integers from 1 to 10
my.integers
length(my.integers) # length of a vector
```


To create a vector in R, the combine function `c()` is used. For
example:
```{r}
numeric.vector <- c(1, 10, 49)
character.vector <- c( "a", "b", "c")
logical.vector <- c(TRUE, FALSE, TRUE)
```



Other useful functions for creating numeric vectors include `seq()` and `rep()`.
```{r}
seq(from=1, to=7, by=3)
rep(1:4, times=2)
```


You can give a name to the elements of a vector with the names() function.
```{r}
names(numeric.vector) <- character.vector
show(numeric.vector)
```


### Operations on vectors {-}

Different types of operations are appropriate for vectors of different type.
Perhaps the most familiar is arithmetic for numeric vectors.
It is important to understand that if you perform arithmetic operations
on two vectors in R, it performs *element-wise* operation:

```{r}
x <- c(6.5, 4.7, 1.2, -3.3)
y <- c(1, 2, 3, 4)

x+y
x/y
```


#### Exercise 2c {-}

1. What happens when we try to do arithmetic with vectors that are of different lengths? Or when we mix a scalar with a vector?
```{r}
x <- c(6.5,  1.2, 4.7, -3.3)
y <- c(1, 2)

```

2. Try applying some mathematical functions to a numeric vector. Use `plot(x,y)` to see the results
```{r}
x <- -10:10
y <-

```


3. Find the `sum`, `mean` and `median` of the data provided.
```{r}
my.data <- c(6.1, 2.7, 1.2, 3.4, 2.0, 8.0, 3.3, 6.1, 1.2)

```

Some other useful functions for vectors: `unique()`,`prod()`,`max()`,`min()`,`range()`,`var()`,`sd()`,`sample()`,`sort()`,`rev()`,`append()`.

4. Try these out below to see how they work.
```{r}

```



### Selection by index {-}

We often need to select elements of numeric vectors (or other data
structures). Such selection can be implemented a number of ways.
The simplest method is numerical indexing using square brackets `[]`.

```{r}
x <- 20:10
x[1]
index <- c(1, 3, 5) # a vector of the indices we want
x[index]
```

Note that indices in R start from `1`, not `0` as in many other languages.

If names have been assigned to the vector elements, we can select by name instead of by numeric position:

```{r}
numeric.vector["c"]
```


### Selection by comparison {-}

It is often useful to do conditional selection. For example, to select only the
positive values from a vector:

```{r}
x <- -5:5
x
x >= 0

index <- x >=0
x[index]
which(index)
```

The vector `index` is logical, containing element-wise Boolean values (`TRUE` or `FALSE`) for the comparison.
The `[]` operator handles a logical vector by selecting only those values that are true.

We use `which(x)` to find the index positions of the `TRUE` elements of a logical vector `x`.



The functions `any()` and `all()` can also be used to test the elements of a vector. These return a single Boolean, e.g.

```{r}
x <- 20:12
x

any(x == 16)
all(x > 12)
```

Another useful operator is `%in%`, which tells you whethe a vector contains a particular object, e.g.

```{r}
x <- 3:9

5 %in% x
0 %in% x
```



#### Exercise 2d {-}
Look at the following expressions. Can you describe what they do?
```{r}
x <- -5:5
index <- x>0
sum(index)
sum(x[index])
```

Note that here the vector `index` is logical, containing the values `FALSE` and `TRUE`.
The `sum()` function has *coerced* these logical values to  numeric ones (reading `FALSE` as `0` and `TRUE` as `1`).
This automatic coercion makes R very flexible, but can also be a source of confusion.

To explicitly coerce objects from one class to another, we can use the `as.*` functions, e.g.
```{r}
index
as.numeric(index)
```


### Combining logical vectors {-}

Logical vectors can be combined element-wise with `!` (logical NOT), `&` (logical AND ) and `∣` (logical OR).

```{r}
x <- 20:10
index.odd <- (x %% 2)
index.odd

index.gt15 <- x > 15
x[index.odd & index.gt15]
```

The vector `index.odd` is actually numeric, containing values `0` and `1`.
The `&` operator has coerced these numeric values to logical ones (reading `0` as `FALSE` and any other number as `TRUE`).


#### Exercise 2e {-}

Which elements of `x` have values between 11 and 18 and are divisible by 3? Use a logical expression.
```{r}

```



### Vector transformations {-}

Having selected a subset of indices, we can make conditional assignments, e.g.

```{r}
x <- 20:10
x[x < 15] <- -1 # assign elements less than 15 to be -1

x
```



#### Exercise 2f {-}

You have started a business selling flapjacks outside the underground station.
The vectors below show your income and expenses for the first five days of trading.

Create the following variables:

1. `profit`, your daily profit (a vector).
2. `profit.total`, your overall profit at the end of the week (a scalar).
3. `profitable.days`, the names of the days on which you made a positive profit (a vector).

```{r}
days <- c("Monday", "Tuesday", "Wednesday","Thursday", "Friday")
income <- c(140, 101, 25, 42, 120)
expenses <- c(220, 23, 0, 51, 34)

```

4. For tax reasons, you decide to donate any daily profit over £50 to charity.
Calculate your daily donations (a vector).

```{r}

```




## Matrices

In R, a matrix is a two-dimensional collection of elements of the
same data type (numeric, character, or logical) arranged into a
fixed number of rows and columns.

A matrix is constructed from a vector by filling positions by column or by row:
```{r}
A <- matrix(1:6, nrow=3) # default filling is by column
A
dim(A) # matrix dimensions (rows first)
```

```{r}
B <- matrix(1:6, byrow=TRUE, nrow=3) # switch to filling by row
B
dim(B) # matrix dimensions (rows first)
```


The functions `colnames()` and `rownames()` are used to add names for the the columns and the rows of a matrix.
```{r}
colnames(A) <- c("red", "green")
rownames(A) <- c("apple", "pepper", "tomato")
A
```

The functions `rowSums()` and `colSums()` conveniently calculate the totals for each row and column of a matrix.
```{r}
colSums(A)
rowSums(A)
```


A diagonal matrix can be constructed using the function `diag()`, e.g.
```{r}
diag(4)
diag(4, x=1:4)
```


Selection (and modification) of elements of a matrix uses the `[]` operator, but with two indices:
```{r}
A[2,2] # single element
A[,2] # single column
A[2,] # single row
```

If only one index is provided, an element is returned by treating the matrix as a vector, read column-wise.
```{r}
A[3]
B[3]
```


The functions `cbind()` and `rbind()` are available for concatenating matrices (or vectors) by row (row bind) or column (column bind).
The dimensions must conform.


#### Exercise 2g {-}

Use `cbind()` and `rbind()` to combine `A` and `B`. What are the dimensions of the resulting matrices?
```{r}

```



### Matrix Transformations {-}

As with vectors, element-wise arithmetic can be applied to matrices.
```{r}
1 + A / 2
```

Note that `*` operates element-wise. For matrix multiplication, we need to use `%*%`
```{r}
C <- matrix(1:4,nrow=2)

C
C * C
C %*% C
```


Many functions can also be applied element-wise to matrices, e.g.
```{r}
log(A)
```

Some other functions taking matrix arguments:

- `nrow(A)`
- `ncol(A)`
- `det(A)` (the determinant of `A`)
- `t(A)` (the transpose of `A`)
- `solve(A,B)` returns `x` such that `A %*% x == B`

If `A` is invertible, `solve(A)` returns the matrix inverse of `A`:
```{r}
invC <- solve(C)
invC
C %*% invC
```


